---
---
title: "Alpha Diversity Analysis"
author: "Nidhi Vijayan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true          # adds a floating table of contents
    toc_depth: 3       # number of heading levels to include
    theme: united      # optional theme (e.g., united, cerulean, cosmo)
    highlight: tango   # syntax highlighting style
---

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Alpha Diversity of mice based on age, sex, and cohort

Based on the below results, I determined that sex difference in alpha diversity were minimal. The alpha diversity and age do not have a linear relationship. Increase in diversity was observed after 5 months old, with a moderate drop after 19 months. The effect was more in "Observed" diversity index, more so than Shannon. 

```{r}
## alpha diversity ====
library(lme4)
library(lmerTest)
library(knitr)
library(dplyr)
library(phyloseq)
library(vegan)
library(emmeans)
library(tibble)
library(reshape2)
library(tidyverse)
library(qiime2R)
library(ggpubr)
library(ggplot2)

theme_set(theme_bw())

load("qiime_f.RData")  
metadata_df<-as.data.frame(sample_data(qiime_f))
metadata_df$SampleID<-rownames(metadata_df)
metadata_df<-as.matrix(metadata_df)

qiime_f_rare<-rarefy_even_depth(qiime_f,sample.size = min(sample_sums(qiime_f)))

alpha.div.df <- estimate_richness(
  qiime_f_rare, 
  measures=c("Shannon", "Observed")) %>% 
  rownames_to_column("SampleID") %>% 
  merge(metadata_df, by="SampleID")
View(alpha.div.df)
alpha.div.long.df <- alpha.div.df %>% 
  pivot_longer(c(Shannon, Observed), names_to="metric")

alpha.div.long.df$month_n<-as.numeric(alpha.div.long.df$month_n)

# summarize across sexes
alpha.div.plot.df <- alpha.div.long.df %>% 
  dplyr::filter(metric %in% c("Shannon", "Observed")) %>% 
  dplyr::filter(month_n <= 40) %>% # omit the few 46 month samples
  
  # compute mean and SD across diets (separately for each metric)
  group_by(metric, sex, month_n, mousegroup.x) %>% 
  summarise(n=n(),
            mean = mean(value),
            sd = sd(value),
            sem = sd(value)/sqrt(n()), .groups="drop") %>% 
  
  # make sure we have at least 3 observations at this age combination
  dplyr::filter(n >= 2) #%>% 

#plot Fig 2 alpha ====
alpha.div.plot.df$month_n <- as.numeric(alpha.div.plot.df$month_n)
alpha.div.plot.df <- alpha.div.plot.df %>%
  arrange(metric, sex, month_n)
```

## Alpha diversity plot

```{r alpha-plot, fig.width=8, fig.height=5, echo=TRUE}
alpha.div.plot.df %>%
  ggplot(aes(
    x = month_n, y = mean,
    ymin = mean - sem, ymax = mean + sem,
    color = sex, group = interaction(sex, metric)
  )) +
  geom_path(alpha = 0.7) +
  geom_pointrange(alpha = 0.7) +
  facet_wrap(~ metric + mousegroup.x, scales = "free", nrow = 2) +
  scale_color_manual(values = c("deeppink", "blue")) +
  labs(
    x = "Age (months)",
    y = "",
    title = "Alpha diversity ASV method"
  ) +
  theme_bw()

```

## LMER for alpha

```{r}
df_wide <- alpha.div.long.df %>%
  pivot_wider(names_from = metric, values_from = value)

```
## Q1: Are there differences between sex and month, and if so, when?
## Q2: Which months differ from baseline in males and females separately?
```{r}

model1 <- lmer(Shannon ~ sex * month + (1 | mousegroup.x) + (1 | mouse), data = df_wide)
kable(anova(model1, type = 3), caption = "ANOVA for model1 with all cohort")
```

```{r}
kable(summary(model1)$coefficients, digits = 4)

```

### Interpretation of the above table

In the above, F M03 is reference. sexM result suggests at M03, Male sig. higher than female at M03 (by 0.1515). Females at M07, M19, N22 were sig different from M03. The sex difference in M07(p=.004), M09 (p=0.03), M23(p=0.0122) is significantly different from the sex difference in M03 .

### ANOVA for model1 (Shannon with all cohorts)

|           |    Sum Sq |   Mean Sq | NumDF |    DenDF |   F value |   Pr(\>F) |
|:----------|----------:|----------:|------:|---------:|----------:|----------:|
| sex       | 0.5794963 | 0.5794963 |     1 | 102.2983 | 10.870027 | 0.0013445 |
| month     | 1.8934174 | 0.1352441 |    14 | 474.3694 |  2.536870 | 0.0016396 |
| sex:month | 1.2073909 | 0.0862422 |    14 | 559.5593 |  1.617707 | 0.0700462 |

```{r}
# Estimated marginal means by Sex within Month for Shannon
emm_sex_within_month <- emmeans(model1, ~ sex | month)
pairs_sex_within_month <- contrast(emm_sex_within_month, method = "pairwise", adjust = "holm")
pairs_sex_within_month

emm_month_within_sex <- emmeans(model1, ~ month | sex)
baseline_contrasts <- contrast(emm_month_within_sex, method = "trt.vs.ctrl", ref = 1, by = "sex")

plot(baseline_contrasts)

```

```{r emmeans_sex_within_month, fig.width=5, fig.height=8, echo=TRUE}
plot(emm_sex_within_month, comparisons = TRUE)
```

### Interpretation of fig

Within each month, we are looking at sex differences. Overall, the females have a higher alpha diversity than males.

## Subsetting per cohort

### Cohort G1

```{r}
df_wide_G1<- df_wide %>%
  filter(mousegroup.x== "G1")
View(df_wide_G1)

model1_G1 <- lmer(Shannon ~ sex * month + (1 | mouse), data = df_wide_G1)
summary(model1_G1)

# From the summary result: There was no difference between sexes at M03, or difference with females, or relationship to M-F relative to baseline
```

```{r}
kable(anova(model1_G1, type = 3), caption ="ANOVA for Cohort G1: Shannon")
```

```{r}

model2_G1 <- lmer(Observed ~ sex * month + (1 | mouse), data = df_wide_G1)
```

### cohort G1: Observed

From the below table:

1.  There is no difference between male and females at M03
2.  Females at month M07, M09, M13, M16, and M17 are significantly different from females at M03
3.  Compared to the difference between males and females in M03, there was no sig difference between Males and Females in other months

|               | Estimate | Std. Error |  df | t value | Pr(\>\|t\|) |
|:--------------|---------:|-----------:|----:|--------:|------------:|
| (Intercept)   | 123.1429 |     8.4062 | 122 | 14.6491 |      0.0000 |
| sexM          |   2.3016 |    11.2082 | 122 |  0.2053 |      0.8376 |
| monthM05      |   5.5238 |    12.3736 | 122 |  0.4464 |      0.6561 |
| monthM07      |  45.4286 |    11.8881 | 122 |  3.8213 |      0.0002 |
| monthM09      |  45.8571 |    11.5106 | 122 |  3.9839 |      0.0001 |
| monthM11      |  16.9683 |    11.2082 | 122 |  1.5139 |      0.1326 |
| monthM13      |  46.1905 |    12.3736 | 122 |  3.7330 |      0.0003 |
| monthM15      |  32.1071 |    11.5106 | 122 |  2.7893 |      0.0061 |
| monthM17      |  28.6349 |    11.2082 | 122 |  2.5548 |      0.0119 |
| monthM19      |   7.0238 |    12.3736 | 122 |  0.5676 |      0.5713 |
| monthM22      | -14.4762 |    15.3475 | 122 | -0.9432 |      0.3474 |
| sexM:monthM05 |  -8.7683 |    17.5213 | 122 | -0.5004 |      0.6177 |
| sexM:monthM07 |  -6.4286 |    15.8509 | 122 | -0.4056 |      0.6858 |
| sexM:monthM09 | -15.0516 |    15.7888 | 122 | -0.9533 |      0.3423 |
| sexM:monthM11 |   8.2123 |    15.5697 | 122 |  0.5275 |      0.5988 |
| sexM:monthM13 |  -2.4921 |    16.6952 | 122 | -0.1493 |      0.8816 |
| sexM:monthM15 | -12.5516 |    16.0661 | 122 | -0.7812 |      0.4362 |
| sexM:monthM17 | -15.0794 |    16.2181 | 122 | -0.9298 |      0.3543 |
| sexM:monthM19 |  28.5317 |    16.4285 | 122 |  1.7367 |      0.0850 |
| sexM:monthM22 |  17.1984 |    19.3119 | 122 |  0.8906 |      0.3749 |

### ANOVA of ModelG1: Observed

|           |      Sum Sq |    Mean Sq | NumDF | DenDF |   F value |         Pr(\>F) |
|:----------|------------:|-----------:|------:|------:|----------:|----------------:|
| sex       |    91.47856 |   91.47856 |     1 |   122 | 0.1849369 |       0.6679231 |
| month     | 36277.26342 | 4030.80705 |     9 |   122 | 8.1488478 | 2.09e-09 \*\*\* |
| sex:month |  6297.16741 |  699.68527 |     9 |   122 | 1.4145130 |       0.1890481 |



```{r}
# Estimated marginal means by Sex within Month for Shannon
emm_sex_within_month_G1 <- emmeans(model1_G1, ~ sex | month)
pairs_sex_within_month_G1 <- contrast(emm_sex_within_month, method = "pairwise", adjust = "holm")

emm_month_within_sex_G1 <- emmeans(model1_G1, ~ month | sex)
baseline_contrasts_G1 <- contrast(emm_month_within_sex_G1, method = "trt.vs.ctrl", ref = 1, by = "sex")

plot(baseline_contrasts_G1)
```

```{r emmeans_sex_within_month_G1, fig.width=5, fig.height=8, echo=TRUE}
plot(emm_sex_within_month_G1, comparisons = TRUE)
```


### Cohort C

```{r}
df_wide_C<- df_wide %>%
  filter(mousegroup.x== "C")
model1_C <- lmer(Shannon ~ sex * month + (1 | mouse), data = df_wide_C)

```

### Summary Cohort C: Shannon

From the below table:

1.  There is sig. difference between male and females at M03 (p=0.0005)
2.  Females shannon diversity relatively stable with age
3.  Compared to the difference between males and females in M03, there was sig difference (decreasing) between Males and Females in months M12 (p= 0.0347), M15 (p=0.0320), and M23 (p=0.0006)

|               | Estimate | Std. Error |       df | t value | Pr(\>\|t\|) |
|:--------------|---------:|-----------:|---------:|--------:|------------:|
| (Intercept)   |   3.8023 |     0.0608 | 219.0609 | 62.5293 |      0.0000 |
| sexM          |   0.2902 |     0.0821 | 218.7018 |  3.5329 |      0.0005 |
| monthM06      |   0.0473 |     0.0810 | 193.8009 |  0.5838 |      0.5600 |
| monthM09      |   0.0858 |     0.0792 | 201.6122 |  1.0829 |      0.2802 |
| monthM12      |   0.1241 |     0.0838 | 196.9755 |  1.4807 |      0.1403 |
| monthM15      |   0.0797 |     0.0810 | 193.8009 |  0.9837 |      0.3265 |
| monthM18      |   0.0662 |     0.0891 | 193.1763 |  0.7423 |      0.4588 |
| monthM21      |   0.0266 |     0.0822 | 192.6006 |  0.3238 |      0.7464 |
| monthM23      |   0.1353 |     0.0798 | 192.4630 |  1.6951 |      0.0917 |
| sexM:monthM06 |  -0.1082 |     0.1128 | 195.5182 | -0.9591 |      0.3387 |
| sexM:monthM09 |  -0.1615 |     0.1170 | 202.4634 | -1.3803 |      0.1690 |
| sexM:monthM12 |  -0.2441 |     0.1148 | 196.0534 | -2.1268 |      0.0347 |
| sexM:monthM15 |  -0.2437 |     0.1128 | 195.9673 | -2.1600 |      0.0320 |
| sexM:monthM18 |  -0.2234 |     0.1210 | 196.4263 | -1.8459 |      0.0664 |
| sexM:monthM21 |  -0.2252 |     0.1161 | 198.1406 | -1.9398 |      0.0538 |
| sexM:monthM23 |  -0.3868 |     0.1101 | 193.5140 | -3.5121 |      0.0006 |

```{r}

kable(anova(model1_C, type = 3), caption = "ANOVA for model1 (Shannon)_C")
```

```{r}
model2_C <- lmer(Observed ~ sex * month + (1 | mouse), data = df_wide_C)
kable(summary(model2_C)$coefficients, digits = 4, caption = "Fixed effects for Observed model (Cohort C)")

```

#### In cohort C, no differences in Observed diversity metric.

```{r}
# Estimated marginal means by Sex within Month for Shannon
emm_sex_within_month_C <- emmeans(model1_C, ~ sex | month)
pairs_sex_within_month_C <- contrast(emm_sex_within_month_C, method = "pairwise", adjust = "holm")

emm_month_within_sex_C <- emmeans(model1_C, ~ month | sex)
baseline_contrasts_C <- contrast(emm_month_within_sex_C, method = "trt.vs.ctrl", ref = 1, by = "sex")

plot(baseline_contrasts_C)
```

```{r emmeans_sex_within_month_C, fig.width=5, fig.height=8, echo=TRUE}
plot(emm_sex_within_month_C, comparisons = TRUE)
```


### For Cohort G2

SHANNON

1.  Sig. diff between M and F in M03
2.  Females are stable
3.  No differences in M and F

|               | Estimate | Std. Error |       df | t value | Pr(\>\|t\|) |
|:--------------|---------:|-----------:|---------:|--------:|------------:|
| (Intercept)   |   3.7939 |     0.1246 | 101.2543 | 30.4592 |      0.0000 |
| sexM          |   0.1223 |     0.1505 |  99.3598 |  0.8126 |      0.4184 |
| monthM05      |  -0.2431 |     0.1490 |  92.9471 | -1.6313 |      0.1062 |
| monthM07      |   0.3279 |     0.1489 |  92.7019 |  2.2017 |      0.0302 |
| monthM09      |   0.0667 |     0.1499 |  94.2028 |  0.4450 |      0.6574 |
| monthM11      |   0.1733 |     0.1713 |  97.3980 |  1.0116 |      0.3142 |
| monthM13      |   0.0261 |     0.1600 |  94.1260 |  0.1631 |      0.8708 |
| monthM15      |   0.1091 |     0.1847 |  96.9481 |  0.5908 |      0.5560 |
| monthM17      |  -0.0433 |     0.1544 |  94.8884 | -0.2803 |      0.7799 |
| monthM19      |   0.3025 |     0.1544 |  94.8884 |  1.9587 |      0.0531 |
| monthM21      |  -0.0113 |     0.1825 |  94.0437 | -0.0618 |      0.9509 |
| monthM23      |  -0.0786 |     0.1600 |  94.2136 | -0.4913 |      0.6244 |
| sexM:monthM05 |   0.2209 |     0.1907 |  92.3040 |  1.1579 |      0.2499 |
| sexM:monthM07 |  -0.2129 |     0.1995 |  92.9569 | -1.0674 |      0.2886 |
| sexM:monthM09 |  -0.1397 |     0.1915 |  93.3516 | -0.7293 |      0.4677 |
| sexM:monthM11 |  -0.1075 |     0.2087 |  95.8024 | -0.5149 |      0.6078 |
| sexM:monthM13 |   0.0446 |     0.1995 |  93.2802 |  0.2236 |      0.8236 |
| sexM:monthM15 |  -0.1136 |     0.2231 |  95.7534 | -0.5091 |      0.6119 |
| sexM:monthM17 |   0.0029 |     0.2036 |  94.1650 |  0.0142 |      0.9887 |
| sexM:monthM19 |  -0.2752 |     0.2108 |  94.4461 | -1.3056 |      0.1949 |
| sexM:monthM21 |  -0.0380 |     0.2256 |  93.7753 | -0.1683 |      0.8667 |
| sexM:monthM23 |  -0.0581 |     0.2079 |  93.9131 | -0.2794 |      0.7805 |

: Fixed effects for Observed model (Cohort G2)

OBSERVED For G2

1.  No diff at baseline
2.  M07 and M19 females are diff from M03
3.  Marginal diff between M and F in M19

|               | Estimate | Std. Error |       df | t value | Pr(\>\|t\|) |
|:--------------|---------:|-----------:|---------:|--------:|------------:|
| (Intercept)   | 116.9246 |    11.8680 | 100.9588 |  9.8521 |      0.0000 |
| sexM          |  13.2977 |    14.3580 |  98.5733 |  0.9261 |      0.3566 |
| monthM05      |  11.7280 |    14.1262 |  92.5827 |  0.8302 |      0.4085 |
| monthM07      |  45.4176 |    14.1182 |  92.3351 |  3.2170 |      0.0018 |
| monthM09      |  24.7788 |    14.2162 |  93.7734 |  1.7430 |      0.0846 |
| monthM11      |  13.0188 |    16.2621 |  96.9500 |  0.8006 |      0.4253 |
| monthM13      |   3.5086 |    15.1743 |  93.7729 |  0.2312 |      0.8176 |
| monthM15      |  16.5746 |    17.5303 |  96.4958 |  0.9455 |      0.3468 |
| monthM17      |  17.7518 |    14.6462 |  94.5244 |  1.2120 |      0.2285 |
| monthM19      |  56.7518 |    14.6462 |  94.5244 |  3.8748 |      0.0002 |
| monthM21      | -10.5578 |    17.3059 |  93.6354 | -0.6101 |      0.5433 |
| monthM23      |  17.0319 |    15.1767 |  93.8676 |  1.1222 |      0.2646 |
| sexM:monthM05 |  15.8732 |    18.0796 |  91.9626 |  0.8780 |      0.3823 |
| sexM:monthM07 | -24.2694 |    18.9094 |  92.6070 | -1.2835 |      0.2025 |
| sexM:monthM09 | -13.9993 |    18.1584 |  92.9841 | -0.7710 |      0.4427 |
| sexM:monthM11 |   2.7607 |    19.8010 |  95.3951 |  0.1394 |      0.8894 |
| sexM:monthM13 |  15.7810 |    18.9142 |  92.9558 |  0.8343 |      0.4062 |
| sexM:monthM15 |  -9.2275 |    21.1623 |  95.3389 | -0.4360 |      0.6638 |
| sexM:monthM17 |  -8.5881 |    19.3048 |  93.8117 | -0.4449 |      0.6574 |
| sexM:monthM19 | -37.3228 |    19.9866 |  94.0765 | -1.8674 |      0.0650 |
| sexM:monthM21 |   5.2741 |    21.3946 |  93.3928 |  0.2465 |      0.8058 |
| sexM:monthM23 |  -8.5674 |    19.7165 |  93.5724 | -0.4345 |      0.6649 |

: Fixed effects for Observed model (Cohort G2)

```{r}
# Estimated marginal means by Sex within Month for Shannon
df_wide_G2<- df_wide %>%
  filter(mousegroup.x== "G2")
  
model1_G2 <- lmer(Shannon ~ sex * month + (1 | mouse), data = df_wide_G2)
emm_sex_within_month_G2 <- emmeans(model1_G2, ~ sex | month)
pairs_sex_within_month_G2 <- contrast(emm_sex_within_month_G2, method = "pairwise", adjust = "holm")

emm_month_within_sex_G2 <- emmeans(model1_G2, ~ month | sex)
baseline_contrasts_G2 <- contrast(emm_month_within_sex_G2, method = "trt.vs.ctrl", ref = 1, by = "sex")

plot(baseline_contrasts_G2)
```

```{r emmeans_sex_within_month_G2, fig.width=5, fig.height=8, echo=TRUE}
plot(emm_sex_within_month_G2, comparisons = TRUE)
```


### Cohort G3

OBSERVED

M11 females diff from M03 females

|               | Estimate | Std. Error |  df | t value | Pr(\>\|t\|) |
|:--------------|---------:|-----------:|----:|--------:|------------:|
| (Intercept)   | 151.6000 |     9.3928 | 100 | 16.1401 |      0.0000 |
| sexM          |   4.4000 |    11.5037 | 100 |  0.3825 |      0.7029 |
| monthM05      |   4.6000 |    13.2834 | 100 |  0.3463 |      0.7298 |
| monthM07      |  -2.2667 |    12.7179 | 100 | -0.1782 |      0.8589 |
| monthM09      |   8.4000 |    15.3383 | 100 |  0.5476 |      0.5852 |
| monthM11      |  33.9000 |    14.0892 | 100 |  2.4061 |      0.0180 |
| monthM13      |  11.1500 |    14.0892 | 100 |  0.7914 |      0.4306 |
| monthM15      |   8.0667 |    15.3383 | 100 |  0.5259 |      0.6001 |
| monthM17      | -29.6000 |    23.0075 | 100 | -1.2865 |      0.2012 |
| monthM19      |  -5.8000 |    13.2834 | 100 | -0.4366 |      0.6633 |
| monthM22      | -20.7667 |    12.7179 | 100 | -1.6329 |      0.1056 |
| monthM23      | -14.0000 |    22.0280 | 100 | -0.6356 |      0.5265 |
| sexM:monthM05 |  12.5111 |    16.4187 | 100 |  0.7620 |      0.4479 |
| sexM:monthM07 |   2.9667 |    15.8104 | 100 |  0.1876 |      0.8515 |
| sexM:monthM09 | -44.2571 |    18.5039 | 100 | -2.3918 |      0.0186 |
| sexM:monthM11 | -25.0000 |    16.9331 | 100 | -1.4764 |      0.1430 |
| sexM:monthM13 |  13.3500 |    17.7802 | 100 |  0.7508 |      0.4545 |
| sexM:monthM15 |   5.3778 |    18.1215 | 100 |  0.2968 |      0.7673 |
| sexM:monthM17 |  24.1000 |    25.4357 | 100 |  0.9475 |      0.3457 |
| sexM:monthM19 | -13.6000 |    17.5723 | 100 | -0.7739 |      0.4408 |
| sexM:monthM22 |  13.2667 |    16.7146 | 100 |  0.7937 |      0.4292 |

SHANNON

|               | Estimate | Std. Error |  df | t value |     Pr() |
|:--------------|---------:|-----------:|----:|--------:|---------:|
| (Intercept)   |   3.9141 |     0.1000 | 100 | 39.1539 |   0.0000 |
| sexM          |   0.0736 |     0.1224 | 100 |  0.6010 |   0.5492 |
| monthM05      |   0.0475 |     0.1414 | 100 |  0.3358 |   0.7377 |
| monthM07      |   0.0356 |     0.1354 | 100 |  0.2631 |   0.7930 |
| monthM09      |   0.1372 |     0.1632 | 100 |  0.8405 |   0.4026 |
| monthM11      |   0.1876 |     0.1499 | 100 |  1.2514 |   0.2137 |
| monthM13      |   0.1749 |     0.1499 | 100 |  1.1666 |   0.2462 |
| monthM15      |   0.3460 |     0.1632 | 100 |  2.1197 | 0.0365\* |
| monthM17      |   0.2787 |     0.2449 | 100 |  1.1383 |   0.2577 |
| monthM19      |  -0.0303 |     0.1414 | 100 | -0.2143 |   0.8307 |
| monthM22      |  -0.3130 |     0.1354 | 100 | -2.3122 | 0.0228\* |
| sexM:monthM05 |  -0.0310 |     0.1747 | 100 | -0.1777 |   0.8594 |
| sexM:monthM07 |  -0.0396 |     0.1683 | 100 | -0.2354 |   0.8144 |
| sexM:monthM09 |  -0.2277 |     0.1969 | 100 | -1.1564 |   0.2503 |
| sexM:monthM11 |  -0.0360 |     0.1802 | 100 | -0.2000 |   0.8419 |
| sexM:monthM13 |   0.0120 |     0.1892 | 100 |  0.0635 |   0.9495 |
| sexM:monthM15 |  -0.1370 |     0.1929 | 100 | -0.7106 |   0.4790 |
| sexM:monthM17 |  -0.3000 |     0.2707 | 100 | -1.1080 |   0.2705 |
| sexM:monthM19 |  -0.0251 |     0.1870 | 100 | -0.1342 |   0.8935 |
| sexM:monthM22 |   0.1546 |     0.1779 | 100 |  0.8691 |   0.3869 |

```{r}
# Estimated marginal means by Sex within Month for Shannon
df_wide_G3<- df_wide %>%
  filter(mousegroup.x== "G3")
  
model1_G3 <- lmer(Shannon ~ sex * month + (1 | mouse), data = df_wide_G3)
emm_sex_within_month_G3 <- emmeans(model1_G3, ~ sex | month)
pairs_sex_within_month_G3 <- contrast(emm_sex_within_month_G3, method = "pairwise", adjust = "holm")

emm_month_within_sex_G3 <- emmeans(model1_G3, ~ month | sex)
baseline_contrasts_G3 <- contrast(emm_month_within_sex_G3, method = "trt.vs.ctrl", ref = 1, by = "sex")

plot(baseline_contrasts_G3)
```

```{r emmeans_sex_within_month_G3, fig.width=5, fig.height=8, echo=TRUE}
plot(emm_sex_within_month_G3, comparisons = TRUE)
```

## Alpha Diversity of mice based on frailty

To determine potential relationship between frailty scores and survival, I filtered the dataset to the last time point of collection

```{r}
metadata<-read.delim("Young_aging_metadata_v4.txt")

# keeping only the last one per mouse
metadata_sub <- metadata %>%
  group_by(mouse) %>%
  arrange(month_n, .by_group = TRUE) %>%  # ensure chronological order
  slice_tail(n = 1) %>%                 
  ungroup()
```
This results in 186 mice

```{r}
#Remove samples with no frailty information
metadata_sub_f<-metadata_sub %>%
  filter(frailty_physical %in% c("PF","NF","F"))

write.table(metadata_sub_f,"metadata_frailty.txt",sep="\t", quote = FALSE,
            row.names = FALSE)

qiime_frail<-qza_to_phyloseq(features="Aging-table-dada2.qza",
                       tree="rooted-tree-merged.qza",
                       metadata="metadata_frailty.txt", taxonomy = "Aging-taxonomy-weighted_silva138-1.qza")

sample_reads <- sample_sums(qiime_frail)
sorted_reads <- sort(sample_reads)
print(sorted_reads)

qiime.rare.frail<- rarefy_even_depth(qiime_frail,sample.size = min(sample_sums(qiime_frail)))

```
### Alpha diversity 
#### I found no association with frailty scores and microbiota
```{r}
metadata_df_frail<-as.data.frame(sample_data( qiime.rare.frail))
metadata_df_frail$SampleID<-rownames(metadata_df_frail)
metadata_df_frail<-as.matrix(metadata_df_frail)

alpha.div.df.frail <- estimate_richness(
  qiime.rare.frail, 
  measures=c("Shannon", "Observed")) %>% 
  rownames_to_column("SampleID") %>% 
  merge(metadata_df_frail, by="SampleID")

alpha.div.df.frail.long <- alpha.div.df.frail %>% 
  pivot_longer(c(Shannon, Observed), names_to="metric")

alpha.div.plot.df.frail <- alpha.div.df.frail.long %>%
  arrange(metric, frailty_physical)

a_my_comparisons_frail <-list(c("F","NF"),c("NF","PF"),c("PF","F"))
symnum.args <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns"))

alpha.div.df.frail.long%>% 
  ggplot(aes(x=frailty_physical, y=value, color=frailty_physical))+
geom_boxplot()+facet_wrap(~metric, scales="free")+
  theme_set(theme_bw())+
  theme(axis.text.x=element_text(angle=30))+
  stat_compare_means(comparisons = a_my_comparisons_frail, symnum.args = symnum.args, method = "t.test")

#No. sig difference found here
  
```
```{r}
df_wide_frail<- alpha.div.df.frail.long%>%
  pivot_wider(names_from = metric, values_from = value)
model_shannon_frail <- lmer(Shannon ~ frailty_physical + (1 | mousegroup.x), data = df_wide_frail)
kable(summary(model_shannon_frail)$coefficients, digits = 4)
```
```{r}
df_wide_frail$howlette_frailty<-as.numeric(df_wide_frail$howlette_frailty)
model_shannon_hf <- lmer(Shannon ~ howlette_frailty + (1 | mousegroup.x), data = df_wide_frail)
kable(summary(model_shannon_hf)$coefficients, digits = 4)
```




