---
title: "Statistical Analysis of beta diversity"
author: "Nidhi Vijayan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: flatly
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PERMANOVA on Bray-Curtis distance

```{r bray-dist, message=FALSE, warning=FALSE}
library(phyloseq)
library(vegan)
library(pairwiseAdonis)
library(dplyr)
library(tibble)
library(ggplot2)
library(tidyr)
library(tinytex)
theme_set(theme_bw())

load("qiime_norm.RData")
load("qiime_f.RData")

samples_df<-as.matrix(sample_data(qiime_norm))
samples_df<-as.data.frame(samples_df)
samples_df$sex_month <-paste(samples_df$sex,samples_df$month,sep="_")
bray_dist_matrix<-phyloseq::distance(qiime_norm, method ="bray")
```

```{r adonis1}
adonis2(bray_dist_matrix ~ mousegroup.x + month+sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="terms")

```
```{r}
bd1<-betadisper(bray_dist_matrix, samples_df$mousegroup.x)
anova(bd1)
```

```{r adonis2}
adonis2(bray_dist_matrix ~ mousegroup.x + month*sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="terms")

```

```{r adonis3}
adonis2(bray_dist_matrix ~ mousegroup.x + month+sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="margin")

```
```{r adonis4}
adonis2(bray_dist_matrix ~ mouse+month+sex, data = samples_df, permutations = 999, by="margin")

```
```{r}
adonis2(bray_dist_matrix ~ mouse+mousegroup.x+month+sex, data = samples_df, permutations = 999, by="margin")
```

```{r unifrac, message=FALSE, warning=FALSE}
wuf_dist_matrix<-phyloseq::distance(qiime_norm, method ="unifrac",weighted=TRUE)
```

```{r adonis1_wuf}
adonis2(wuf_dist_matrix ~ mousegroup.x + month+sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="terms")
```
```{r jacard, message=FALSE, warning=FALSE}
jac_dist_matrix<-phyloseq::distance(qiime_norm, method ="jaccard")
```

```{r adonis1_jac}
adonis2(jac_dist_matrix ~ mousegroup.x + month+sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="terms")
```
```{r adonis1_jac_int}
adonis2(jac_dist_matrix ~ mousegroup.x + month*sex, data = samples_df, permutations = 999, strata = samples_df$mouse, by="terms")
```

```{r pair-adonis}
pair_results<-pairwise.adonis(bray_dist_matrix, samples_df$sex_month)
pairwise_long <- pair_results %>%
  separate(pairs, into = c("Group1", "Group2"), sep = " vs ") %>%
  mutate(sig_label = ifelse(p.adjusted < 0.05, "*", "ns"))

#no relationship was significant
ggplot(pairwise_long, aes(x = Group1, y = Group2, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sig_label), color = "black", size = 0.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Pairwise PERMANOVA (Bray–Curtis)",
       fill = "R²") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
pair_results_time<-pairwise.adonis(bray_dist_matrix, samples_df$month)
pair_results_time
```

### PERMANOVA by sex at each month

```{r}
adonis_by_group <- function(physeq_obj, group_var = "month", formula = "sex", dist_method = "bray") {
  
  sample_df <- data.frame(sample_data(physeq_obj))
  if(!group_var %in% colnames(sample_df)) stop(paste("Variable", group_var, "not found in sample_data"))
  
  groups <- unique(sample_df[[group_var]])
  
  results_list <- lapply(groups, function(g) {
    # Identify sample names for this group
    keep_samples <- rownames(sample_df[sample_df[[group_var]] == g, ])
    
    # Subset using sample_names()
    phy_sub <- prune_samples(keep_samples, physeq_obj)
    
    # Normalize
    phy_norm <- transform_sample_counts(phy_sub, function(x) x / sum(x))
    
    # Distance
    dist_mat <- phyloseq::distance(phy_norm, method = dist_method)
    
    # Metadata
    df <- data.frame(sample_data(phy_sub))
    
    # ✅ Check for single-level variable (goes HERE)
    formula_var <- all.vars(as.formula(paste("~", formula)))[1]
    if (length(unique(df[[formula_var]])) < 2) {
      message(paste("Skipping", g, "because", formula_var, "has only one level"))
      return(NULL)
    }
    
    # Run adonis2 safely
    a <- adonis2(as.formula(paste("dist_mat ~", formula)), data = df, by = "margin")
    
    # Extract R2 and p-values
    data.frame(
      group = g,
      term = rownames(a),
      R2 = a$R2,
      p.value = a$`Pr(>F)`
    )
  })
  
  results_df <- dplyr::bind_rows(results_list)
  return(results_df)
}



adonis_results <- adonis_by_group(qiime_norm,"month_n", formula = "mousegroup.x + sex")
adonis_results<- adonis_results %>%
  filter(term== c("sex"))

```

```{r}
adonis_results_cohort <- adonis_by_group(qiime_norm,"month_n", formula = "mousegroup.x")
adonis_results_cohort<- adonis_results_cohort %>%
  filter(term== c("mousegroup.x"))

```

### PERMANOVA summary of sex diff by month

```{r plot adonis}
ggplot(adonis_results, aes(x=group, y=R2, color=p.value)) +
  geom_point(alpha=0.7, size=3)+
  #geom_segment( aes(xend=group,yend=0)) +
  scale_color_gradient(low="black", high="#218d96")+
  theme(axis.text=element_text(size=12))+
  geom_smooth()
```

### PERMANOVA summary of cohort diff by month

```{r plot adonis2}
ggplot(adonis_results_cohort, aes(x=group, y=R2, color=p.value)) +
  geom_point(alpha=0.7, size=3)+
  #geom_segment( aes(xend=group,yend=0)) +
  scale_color_gradient(low="black", high="#218d96")+
  theme(axis.text=element_text(size=12))+
  geom_smooth()
```
### PERMANOVA figure sex and cohort per month combined
```{r combined}
adonis_summary<-left_join(adonis_results,adonis_results_cohort,by="group")

adonis_summary<-adonis_summary %>%
  rename(p_value.x = p.value.x, p_value.y = p.value.y)
adonis_summary_long<- adonis_summary %>%
  pivot_longer(
    cols = c(term.x, R2.x, p_value.x, term.y, R2.y, p_value.y),
    names_to = c(".value", "set"),      # .value keeps R2/p.value as separatecolumns
    names_sep = "\\."                   # split on "."
  ) %>%
  select(group, term, R2, p_value)

```

```{r combined plot, echo=FALSE}
ggplot(adonis_summary_long, aes(x=group, y=R2, color=term)) +
  geom_point(alpha=0.7, size=3)+
  #geom_segment( aes(xend=group,yend=0)) +
  #scale_color_gradient(low="black", high="#218d96")+
  theme(axis.text=element_text(size=12))+
  geom_smooth()
```


### Subset to females

```{r female}
qiime_female <- subset_samples(qiime_f, sex == "F")
qiime_female<-prune_taxa(taxa_sums(qiime_female) > 0, qiime_female)

#normalize
qiime_female_norm<-transform_sample_counts(qiime_female,function(x) x / sum(x))
#make samplesdf
samples_female_df<-as.matrix(sample_data(qiime_female_norm))
samples_female_df<-as.data.frame(samples_female_df)
#BD distance
bray_dist_matrix_fem<-phyloseq::distance(qiime_female_norm, method ="bray")

```

```{r adonis female}
adonis2(bray_dist_matrix_fem ~ mousegroup.x + month, data = samples_female_df, permutations = 999, strata = samples_female_df$mouse, by="term")

```

### Subset to males

```{r male}
qiime_male <- subset_samples(qiime_f, sex == "M")
qiime_male<-prune_taxa(taxa_sums(qiime_male) > 0, qiime_male)

#normalize
qiime_male_norm<-transform_sample_counts(qiime_male,function(x) x / sum(x))
#make samplesdf
samples_male_df<-as.matrix(sample_data(qiime_male_norm))
samples_male_df<-as.data.frame(samples_male_df)
#BD distance
bray_dist_matrix_male<-phyloseq::distance(qiime_male_norm, method ="bray")

```

```{r adonis male}
adonis2(bray_dist_matrix_male ~ mousegroup.x + month, data = samples_male_df, permutations = 999, strata = samples_male_df$mouse, by="term")

```

The effect of month on females was more than males. This could mean the microbiota in females was more dynamic.

## PERMDISP on Bray-Curtis distance

```{r bd by sex}
anova(vegan::betadisper(bray_dist_matrix, samples_df$sex))
```

```{r bd by month}
anova(vegan::betadisper(bray_dist_matrix, samples_df$month))
```

```{r bd female-month}
anova(vegan::betadisper(bray_dist_matrix_fem, samples_female_df$month))
```

```{r bd male-month}
anova(vegan::betadisper(bray_dist_matrix_male, samples_male_df$month))
```
```{r}
anova(vegan::betadisper(bray_dist_matrix, samples_df$mouse))
```

```{r plot betadisper}
disp_female <- betadisper(bray_dist_matrix_fem, sample_data(qiime_female)$month)
disp_male   <- betadisper(bray_dist_matrix_male, sample_data(qiime_male)$month)

boxplot(disp_female$distances ~ sample_data(qiime_female)$month, main="Females")
boxplot(disp_male$distances ~ sample_data(qiime_male)$month, main="Males")

```

## Which sex changes more with age?
Within-month variability in community composition isn’t equal in both males and females. Some months have tighter clustering (more homogeneous microbiota), while others are more variable.

From both the PERMANOVA and betadisper we can see that:

1.  Females show stronger overall temporal trends (higher R² for month). Also shows significant stochasticity but not to the same extent as the males.
2.  Males show more uneven variability across months (stronger betadispersion), meaning their community variability fluctuates more

## CCA Analysis
```{r cca and plot}
cap_ord <- ordinate(
  physeq = qiime_norm, 
  method = "CAP",
  distance = bray_dist_matrix,
  formula = ~ mousegroup.x + sex + month
)

# CAP plot
cap_plot <- plot_ordination(
  physeq = qiime_norm, 
  ordination = cap_ord, 
  color = "sex", 
  axes = c(1,2)
) + 
  aes(shape = mousegroup.x) + 
  geom_point(aes(colour = sex), alpha = 0.4, size = 4) + 
  geom_point(colour = "grey90", size = 1.5) + 
  scale_color_manual(values = c( "deeppink",  "#3639ff", 
                                 "darkorchid3", "magenta")
  )

# Now add the environmental variables as arrows
arrowmat <- vegan::scores(cap_ord, display = "bp")

# Add labels, make a data.frame
arrowdf <- data.frame(labels = rownames(arrowmat), arrowmat)

# Define the arrow aesthetic mapping
arrow_map <- aes(xend = CAP1, 
                 yend = CAP2, 
                 x = 0, 
                 y = 0, 
                 shape = NULL, 
                 color = NULL, 
                 label = labels)

label_map <- aes(x = 1.3 * CAP1, 
                 y = 1.3 * CAP2, 
                 shape = NULL, 
                 color = NULL, 
                 label = labels)

arrowhead = arrow(length = unit(0.02, "npc"))

# Make a new graphic
cap_plot + 
  geom_segment(
    mapping = arrow_map, 
    size = .5, 
    data = arrowdf, 
    color = "gray", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 4,  
    data = arrowdf, 
    show.legend = FALSE
  )
```
```{r anova cap}
cap <- capscale(bray_dist_matrix ~ mousegroup.x+ month+sex, data = samples_df)
anova(cap, by = "term")
```

```{r}
# Extract table from capscale summary
anova_terms <- anova(cap, by = "terms")
summary_cap <- summary(cap)
total_ss <- summary_cap$tot.chi       # total variance
constrained_ss <- summary_cap$constr.chi
unconstrained_ss <- summary_cap$unconst.chi

# Compute partial R2 manually
term_ss <- anova_terms$"SumOfSqs"
term_R2 <- term_ss / total_ss

R2_df <- data.frame(
  term = rownames(anova_terms),
  R2 = round(term_R2, 4),
  F = round(anova_terms$F, 3),
  p.value = anova_terms$"Pr(>F)"
)

R2_df
```
### Interpreting adonis and cap
Sex and month explains roughly the same proportion of beta-diversity variation regardless of whether we use a distance-based regression (dbRDA) or a permutation-based ANOVA (PERMANOVA). 

## Effect of weight loss and frailty on the microbiota

I took the last time point collected per mice to determine if there was an effect by frailty. Using both adonis and mantel test, I see no association or correlation with bray curtis distance and frailty

I did not consider max weight loss since we have the data for only 6 mice. 

```{r frail}
load("qiime_frail.RData") #See alpha_diversity.Rmd

qiime_frail_norm<-transform_sample_counts(qiime_frail,function(x) x / sum(x))
samples_df_frail<-sample_data(qiime_frail_norm)
samples_df_frail<- as.matrix(samples_df_frail)
samples_df_frail<- as.data.frame(samples_df_frail)
```
```{r adonis frail}
dist_matrix_frail<-phyloseq::distance(qiime_frail_norm, method ="bray")
adonis2(dist_matrix_frail~ mousegroup.x + frailty_physical, data = samples_df_frail, permutations = 999, by="terms")
```
```{r matel corr}
frail_value <- dist(samples_df_frail$howlette_frailty)

mantel_result <- mantel(dist_matrix_frail, frail_value, method = "spearman", permutations = 9999)

mantel_result
```

